---
title: "coding-challenge-01"
author: "Ananya Jogalekar aj463"
format: html
editor: visual
---

## SOCIOL728: Week 01 Coding Challenge

R tutorial referenced is available [here](https://inarwhal.github.io/NetworkAnalysisR-book/ch3-Cross-Sectional-Network-Data-R.html).

## 1. Reading the Data

**First, reading the Attributes dataset.**

The data file is stored in the same directory as this R file.

```{r}
# using stringsAsFactors so that the names will be read as factor variables instead of strings that are difficult to categorize
attributes_data <- read.csv(file = "camp_attributes.csv", stringsAsFactors = T)
```

Checking the object created by reading the data:

```{r}
class(attributes_data)
```

As the tutorial explains, we need to use matrix objects for network operations. So, converting the dataframe object into a matrix object, and adding a useful set of row names and column names to the matrix:

```{r}
# convert to matrix
#attributes_data <- as.matrix(attributes_data)

# add row and col names
#rownames(attributes_data) <- 1:nrow(attributes_data)
#colnames(attributes_data) <- c("ID", "gender", "counselor")

# matrix conversion is possible, but I commented it out because packages used later required a dataframe format

attributes_data
```

**Next, reading the Time dataset.**

```{r}
time_data <- read.csv(file = "dataset01-camp_time1.csv", stringsAsFactors = T)

# according to the tutorial, we can work with a dataframe for now
# we don't need to add row and col names, since the data already has them

time_data
```

**Performing a few checks:**

Testing if the string factoring worked by calling some string indexes and checking if the have levels:

```{r}
time_data$ID[1:5]
```

We do see levels, so I believe the factoring worked well.

Next, checking if a random column is numeric or not:

```{r}
class(time_data[, "JENNIE"])
```

## 2. Constructing a network object with node attributes in the igraph package

```{r}
library(igraph)
```

According to the tutorial, we start by creating a matrix-based network first. For this, we will have to use the `attributes_data` object as that is a matrix.

```{r}
attributes_netbymatrix <- graph_from_adjacency_matrix(adjmatrix = attributes_data, 
                                                 mode = "directed")
```

Given the above error, it seems we can only use this function on the `time_data` since that is a square. Let's convert that into a matrix and use this:

```{r}
# convert to matrix
time_data <- as.matrix(time_data)

# remove index column to avoid non-square matrix error
time_data <- time_data[, -1] # commented out because it was already executed BUT need to execute for future runs!

time_netbymatrix <- graph_from_adjacency_matrix(adjmatrix = time_data, 
                                                 mode = "directed")
```

Okay, it works! Now let's check its class:

```{r}
time_netbymatrix
```

Great, it is an igraph object. It tells us that the size of the network is 18 and number of edges is 54.

Now, we add the gender attribute from the `attributes_data`.

```{r}
time_netbymatrix <- set_vertex_attr(graph = time_netbymatrix, 
                                     name = "gender", 
                                     value = attributes_data$gender)
```

In order to avoid a matrix-induced error here, we had to re-read the attributes data and comment out the matrix operations. Now that it is a dataframe, we should have successfully added gender as an attribute by sourcing it from this attributes dataframe and mapping it onto the igraph network object created in the previous step.

Similarly, we can map `counselor` as an attribute from the `attributes_data`.

```{r}
time_netbymatrix <- set_vertex_attr(graph = time_netbymatrix, 
                                     name = "counselor", 
                                     value = attributes_data$counselor)
```

Re-checking if the network object is intact:

```{r}
time_netbymatrix
```

**We see gender and counselor appear as attributes now!**

## 3. Visualizing the network in the basic plot function in igraph

```{r}
plot(time_netbymatrix)
```

# 4. Constructing a network object with node attributes in the network package

Getting the packages:

```{r}
detach(package:igraph)
library(network)
```

Constructing `time_data` as a network object:

```{r}
time_data_matrix <- as.matrix(time_data)
time_netbymatrix_example2 <- network(x = time_data_matrix, directed = T) 
```

Converting gender and counselor attributes into characters:

```{r}
attributes_data$gender <- as.character(attributes_data$gender)
attributes_data$counselor <- as.character(attributes_data$counselor)
```

Create a list to use as input to the `network()` function:

```{r}
attribute_list <- do.call(list, attributes_data)
attribute_list
```

Creating the network with the `vertex.attr` argument:

```{r}
time_netbymatrix_example2 <- network(x = time_data, directed = T, 
                                      vertex.attr = attribute_list)
time_netbymatrix_example2
```

```{r}
plot(time_netbymatrix_example2)
```

# 5. Moving between igraph and network objects

r in library(intergraph) : there is no package called ‘intergraph’

```{r}
library(intergraph)
```

Specifying the igraph network:

```{r}
network_from_igraph <- asNetwork(time_netbymatrix)
network_from_igraph
```

# 6. Getting some basic network features using igraph

Including each node's indegree/outdegree, the overall density of the network.

## Degree

```{r}
outdeg <- rowSums(time_data_matrix)
outdeg
```

```{r}
indeg <- colSums(time_data_matrix)
indeg
```

Using in-built functions in igraph:

```{r}
detach(package:network)
library(igraph)
```

```{r}
outdeg_igraph <- degree(graph = time_netbymatrix, mode = "out")
indeg_igraph <- degree(graph = time_netbymatrix, mode = "in")
```

```{r}
# MODIFY THIS CODE TO CHECK IF PREVIOUS MANUAL CALCULATIONS MATCH WITH PACKAGE IN-BUILT CALCS
outdeg_igraph
indeg_igraph
```

## Density

```{r}
num_edges <- gsize(time_netbymatrix)
num_edges
```

```{r}
num_nodes <- gorder(time_netbymatrix)
num_nodes
```

```{r}
number_dyads <- (num_nodes * (num_nodes - 1))
```

Using the above calculations to get density:

```{r}
density_time_network <- num_edges / number_dyads
density_time_network
```

Calculation using igraph in-built function:

```{r}
edge_density(time_netbymatrix)
```
